---
title: "RNA-seq analysis of DMD muscle organoids"
bibliography: projects/project_1/references.bib
nocite: '@*'
---

> **Date:** 18/10/2025
>
> **Description:** [Palmieri et al. (2025)](https://www.biorxiv.org/content/10.1101/2023.07.26.550063v4) generated an organoid-like structure from DMD muscle cells (derived from patient iPSCs) co-cultured with fibroblasts. This dataset contains **12 samples** of **3 conditions**: IsoCTR, Untreated (disease), and Treated (disease treated with high dose of AAV gene therapy). Each condition includes **4 biological replicates**. Raw data are retrieved from the NCBI Bioproject database under accession PRJNA1208956.

# From FASTQ files to count table

<details>

<summary>**Retrieve fastq files from database**</summary>

```{bash eval=FALSE, include=TRUE}
# Get table from ENA API
# This table contains the links to all raw fastq files

curl -s "https://www.ebi.ac.uk/ena/portal/api/filereport?accession=PRJNA1208956&result=read_run&fields=run_accession,fastq_ftp" > ena_runs.txt

# Download all files
# Paired-end reads ==> 2 fastq files for each sample

awk -F'\t' 'NR>1 {print $2}' ena_runs.txt | tr ';' '\n' | while read url; do
    wget ftp://$url
done
```

</details>

<details>

<summary>**Quality control of raw reads**</summary>

Raw fastq files are imported into [FastQC](https://www.bioinformatics.babraham.ac.uk/projects/fastqc/) for some quality control checks. In return, we receive an HTML-based report on each fastq file.

```{bash eval=FALSE, include=TRUE}
# module load fastqc
# mkdir -p reads_qc

for sample in fastq_files/*.fastq.gz
do
  base=$(basename $sample .fastq.gz)
  fastqc "fastq_files/$base.fastq.gz" --outdir reads_qc
done
```

Looks like the adapters haven't been cut. We'll use [`fastp`](https://github.com/OpenGene/fastp), an all-in-one fastq file preprocessor, for this job.

```{bash eval=FALSE, include=TRUE}
# module load fastp
# mkdir -p clean_reads

for sample in fastq_files/*_1.fastq.gz
do
    base=$(basename "$sample" _1.fastq.gz)

    fastp \
        -i "fastq_files/${base}_1.fastq.gz" \
        -I "fastq_files/${base}_2.fastq.gz" \
        -o "clean_reads/${base}_1_clean.fastq.gz" \
        -O "clean_reads/${base}_2_clean.fastq.gz" \
        -h "clean_reads/${base}_fastp.html" \
        -j "clean_reads/${base}_fastp.json" \
        -w 12     # run one sample at a time using 12 threads
done
```

| Note: If the job keeps crashing, try to do half the number of samples until sorted BAM, delete fastq files, then proceed the other half.

Check folder size regularly:

```{bash eval=FALSE, include=TRUE}
du -sh /shared/home/minhytran/* | sort -h
```

</details>

<details>

<summary>**Read mapping to reference genome**</summary>

Prior to alignment, we need to download human reference genome (GRCh38/hg38) and index it.

```{bash eval=FALSE, include=TRUE}
# Download reference genome
curl -o ref_genome/GCF_000001405.40_GRCh38.p14_genomic.fna.gz \
https://ftp.ncbi.nlm.nih.gov/genomes/all/GCF/000/001/405/GCF_000001405.40_GRCh38.p14/GCF_000001405.40_GRCh38.p14_genomic.fna.gz

# To decompress fasta file :
# gunzip GCF_000001405.40_GRCh38.p14_genomic.fna.gz


# Build genome indexes :
# module load hisat2
# mkdir -p ref_genome/index_hisat2 
hisat2-build ref_genome/Homo_sapiens.GRCh38.dna.primary_assembly.fa \
ref_genome/index_hisat2/human
```

Now we'll map the (paired) reads to the reference. Remember to use 'clean' fastq files. To avoid crashing the alignment, directly output sorted BAM files (skip SAM and unsorted BAM).

```{bash eval=FALSE, include=TRUE}
# module load hisat2 samtools
# mkdir -p reads_mapping

for sample in clean_reads/SRR3207614*_1_clean.fastq.gz 
do 
  
  base=$(basename "$sample" _1_clean.fastq.gz)
  
  hisat2 -q -p 12 \
  --summary-file "./reads_mapping/summary_${base}.txt" \
  -x "./ref_genome/index_hisat2/human" \
  -1 "./clean_reads/${base}_1_clean.fastq.gz" \
  -2 "./clean_reads/${base}_2_clean.fastq.gz" \
| samtools sort -@ 12 -o ./reads_mapping/${base}.sorted.bam
  
done
```

</details>

<details>

<summary>**Manipulate SAM/BAM file with Samtools**</summary>

-   Convert SAM to BAM file (binary format) → save space (done previously)

-   BAM files need to be *sorted* (done previously) and *indexed*

```{bash eval=FALSE, include=TRUE}
# module load samtools

for sample in reads_mapping/*.sorted.bam
do 
  base=$(basename "$sample" .sorted.bam)
  
  # Index sorted BAM
  samtools index "reads_mapping/${base}.sorted.bam"

done
```

</details>

<details>

<summary>**Generate count table**</summary>

Now we'll pass all the sorted, indexed BAM files through `featureCounts` function (subread package) to generate the count matrix. We also need an annotation file to do this. Parameters `-t "exon" -g gene_id` are to count reads mapped to exon then summarize to gene-level (i.e. many exons belong to single gene id).

```{bash eval=FALSE, include=TRUE}
# Download annotation file (.gtf)
wget ftp://ftp.ensembl.org/pub/release-100/gtf/homo_sapiens/Homo_sapiens.GRCh38.100.gtf.gz
# Unzip (then move to ref_genome folder)
gunzip Homo_sapiens.GRCh38.100.gtf.gz

# module load subread
# Remember -p to tell featureCounts that BAM files contain paired-end reads

featureCounts -a ./ref_genome/Homo_sapiens.GRCh38.100.gtf \
-s 0 -t "exon" -g gene_id \
-p --countReadPairs \
-o ./count_matrix.counts \
./reads_mapping/SRR*.sorted.bam
```

</details>

Now we can read the **count matrix** into R:

```{r read_count_table, echo=TRUE, message=FALSE, warning=FALSE}
# Read count table
count_tab <- read.table(file = "./projects/project_1/dmd-organoid_count_matrix.counts",
                        header = TRUE,
                        row.names = 1,
                        sep = "\t",
                        comment.char = "#")

# Clean column names
colnames(count_tab) <- gsub("^.*\\/", "", 
                         colnames(count_tab))    # to match everything from the start of the string up to and including the last slash /

colnames(count_tab) <- gsub("\\.sorted\\.bam$", "", 
                         colnames(count_tab))    # drop ".sorted.bam"

colnames(count_tab) <- gsub("^\\.\\.reads_mapping\\.", "", 
                            colnames(count_tab)) # from the start, drop "..reads_mapping"

colnames(count_tab)

# Drop annotation columns and keep the counts only
count_tab <- count_tab[, -c(1:5)]
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
knitr::kable(head(count_tab), caption = "Count table: genes as rows, samples as columns") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

And **metadata** file:

```{r read_metadata, echo=TRUE, message=FALSE, warning=FALSE}
library(readxl)
metadata <- read_excel("./projects/project_1/dmd-organoid_metadata.xlsx")
meta <- as.data.frame(metadata[,c(1,3)]) # keep 2 columns: sample and sample_type
meta$sample_type <- factor(meta$sample_type) # sample_type as factor

# sample names have to be rownames
rownames(meta) <- meta$sample
meta$sample <- NULL

# Map row names (sample names) from meta to column names of count table
name_map <- setNames(metadata$sample, metadata$sample_id)
colnames(count_tab) <- name_map[colnames(count_tab)]

# Reorder meta to match the order of columns in count table
meta <- meta[colnames(count_tab), , drop=FALSE]

# Check if it's good to go
all(rownames(meta) == colnames(count_tab))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(meta, caption = "Sample metadata") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

# Differential expression analysis with DESeq2

## Create DESeqDataSet object

DESeq2 expects a raw count matrix as input. The model will internally correct for library size. Remember that `nrow(colData) == ncol(countData)`.

```{r create_deseq2_obj, echo=TRUE, message=FALSE, warning=FALSE}
library(DESeq2)
dds <- DESeqDataSetFromMatrix(
  countData = count_tab, 
  colData = meta,
  design = ~ sample_type)
dds
```

## Filter low-count genes

Here we're keeping only genes that have at least 10 read counts in at least 4 samples. The mean-sd plot suggests that variance is stabilized, meaning genes with low and high mean counts all have similar variances.

```{r gene_filtering, echo=TRUE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
smallestGroupSize <- 4
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds_filt <- dds[keep,]

library(vsn)
vsd <- vst(dds_filt)
meanSdPlot(assay(vsd))
```

## Data quality control by clustering samples

For differential gene expression (DEG) analysis, DESeq2 operates directly on the raw counts. However, for clustering and visualization, we better use *normalized* and *transformed* expression values. DESeq2 uses a normalization method called **median of ratios**.

| Note: We have to run DEG analysis to add normalization factors (since it's an all-in-one function)

```{r QC_sample_clustering, echo=TRUE, message=FALSE, warning=FALSE}
# DESeqDataSet object does not store a matrix of normalized counts by default
# We can to get it by: counts(dds, normalized=TRUE)
dds_filt <- DESeq(dds_filt)

select_top_genes <- order(rowMeans(counts(dds_filt, normalized=TRUE)), # mean of normalized counts of each gene across samples                 
                                     decreasing=TRUE)[1:50] # select top 50 genes with highest mean expression  

df <- as.data.frame(colData(dds_filt)[, "sample_type", drop=FALSE])

# log2 transformation of normalized counts 
# pc = 1 is pseudocount ==> log2(n + 1) 
ntd <- normTransform(dds_filt, f = log2, pc = 1)
```

This is a heatmap of top 50 highly expressed genes. Expression values are presented as normalized, log2-transformed counts and are row-scaled (so that mean expression of each gene across samples is 0). Here we see samples are clustered based on top gene expression.

```{r top_gene_heatmap, echo=TRUE, fig.height=12, fig.width=8, message=FALSE, warning=FALSE}
library("pheatmap")
pheatmap(assay(ntd)[select_top_genes,],          
         scale = "row", # z score     
         cluster_rows=TRUE,           
         show_rownames=TRUE,                    
         cluster_cols=TRUE,           
         annotation_col=df)
```

We can make a heatmap of (euclidean) distances between each pair of samples. With this, we have an overall idea of how samples are similar to each other.

```{r dist_matrix_heatmap, echo=TRUE, fig.height=8, fig.width=8, message=FALSE, warning=FALSE}
sampleDists <- dist(t(assay(vsd)))  

library("RColorBrewer") 

sampleDistMatrix <- as.matrix(sampleDists)
colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255) 

pheatmap(sampleDistMatrix,          
         clustering_distance_rows=sampleDists,    
         clustering_distance_cols=sampleDists,          
         col=colors)
```

Another way to evaluate the batch effects is to plot the principle components of samples (PCA).

```{r pca, echo=TRUE, fig.height=5, fig.width=5, message=FALSE, warning=FALSE}
plotPCA(vsd, intgroup="sample_type", ntop=10000)
```

## DEG analysis

`DESeq()` performs:

-   estimation of size factors

-   estimation of dispersion

-   Negative Binomial GLM fitting and Wald statistics

```{r diff_analysis, echo=TRUE, message=FALSE, warning=FALSE}
# If needed, set 'IsoCTR' as baseline condition :
# dds_filt$sample_type <- relevel(dds_filt$sample_type, ref = "IsoCTR")
# dds_filt <- DESeq(dds_filt)
resultsNames(dds_filt) # names of the individual effects (coefficients)
```

p values are corrected by Benjamin-Hochberg with `alpha = 0.05`. **How many genes were (significantly) up/down-regulated in 'Treated' compared to 'IsoCTR'?**

```{r echo=TRUE}

Treated_vs_IsoCTR <- results(dds_filt,
                             name = "sample_type_Treated_vs_IsoCTR",
                             alpha = 0.05)
summary(Treated_vs_IsoCTR)
```

For the volcano plot, we'll use the full `results` table and highlight the significant hits. Be careful that there are genes where `padj` is missing due to independent filtering inside DESeq2 (prior to multiple testing correction) or `log2FoldChange` is missing if a gene has zero counts across all samples of one condition.

<details>

<summary>**Make a volcano plot of DEGs**</summary>

```{r volcano_plot, echo=TRUE, fig.height=7, fig.width=7, message=FALSE, warning=FALSE}

# Libraries 
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tibble)

# Prepare DEGs data frame
Treated_vs_IsoCTR_DEGs <- Treated_vs_IsoCTR %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "ENSEMBL") %>% 
  # Remove genes where log2FC or padj is missing
  filter(!is.na(log2FoldChange), !is.na(padj))

# Add column specifying Up/Down/NotSig genes
Treated_vs_IsoCTR_DEGs$regulation <- ifelse(
  Treated_vs_IsoCTR_DEGs$padj < 0.05 & 
    Treated_vs_IsoCTR_DEGs$log2FoldChange > 1, "Up",
  ifelse( Treated_vs_IsoCTR_DEGs$padj < 0.05 &
            Treated_vs_IsoCTR_DEGs$log2FoldChange < -1, "Down",
          "NotSig")
  )
  
# Count up/down genes
counts <- Treated_vs_IsoCTR_DEGs %>%
  filter(regulation %in% c("Up","Down")) %>%
  group_by(regulation) %>%
  summarise(n = n(), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = regulation, values_from = n, values_fill = 0)

# Volcano plot
p <- ggplot(data = Treated_vs_IsoCTR_DEGs,
            aes(x = log2FoldChange,
                y = -log10(padj))) +
  geom_point(aes(col = regulation)) +
  
  # Add threshold lines
  geom_hline(yintercept = -log10(0.05), 
             col = "black", 
             linetype = 'dashed') +
  
  geom_vline(xintercept = c(-1, 1),
           col = "black",
           linetype = 'dashed') +
  
  scale_color_manual(values = c("Down"="#00AFBB",
                                "NotSig"="grey",
                                "Up"="#E41A1C"),
                     labels = c("Down"="Downregulated", 
                                "NotSig"="Not significant", 
                                "Up"="Upregulated")) +
  
  labs(x = "log2FC",
       y = "-log10(padj)",
       color = "Regulation", 
       title = "Treated vs IsoCTR") + 
  theme_minimal() +
  
  # Annotate number of up/down genes on the plot 
  annotate("text", x = max(Treated_vs_IsoCTR_DEGs$log2FoldChange, na.rm=TRUE), 
           y = max(-log10(Treated_vs_IsoCTR_DEGs$padj), na.rm=TRUE), 
           label = paste0("Up: ", counts$Up), 
           hjust = 1, vjust = 1, col = "#E41A1C") +
  
  annotate("text", x = min(Treated_vs_IsoCTR_DEGs$log2FoldChange, na.rm=TRUE), 
           y = max(-log10(Treated_vs_IsoCTR_DEGs$padj), na.rm=TRUE), 
           label = paste0("Down: ", counts$Down), 
           hjust = 0, vjust = 1, col = "#00AFBB")
```

</details>

```{r display_volcano_plot, echo=FALSE, message=FALSE, warning=FALSE}
p
```

<details>

<summary>Save DEG list to file</summary>

```{r save_DEGs, echo=TRUE, message=FALSE, warning=FALSE}
# Save DEG list
write.table(Treated_vs_IsoCTR_DEGs,
            file = "./projects/project_1/DEG_List_backup/Treated_vs_IsoCTR_DEGs.txt",
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)
```

</details>

**How many genes were (significantly) up/down-regulated in 'Untreated' compared to 'IsoCTR'?**

```{r echo=TRUE, message=FALSE, warning=FALSE}
Untreated_vs_IsoCTR <- results(dds_filt,
                               name = "sample_type_Untreated_vs_IsoCTR",
                               alpha = 0.05)
summary(Untreated_vs_IsoCTR)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# Remove genes where log2FC or padj is missing
# Make row names a column EMSEMBL

Untreated_vs_IsoCTR_DEGs <- 
  Untreated_vs_IsoCTR %>%
  as.data.frame() %>% 
  filter(!is.na(log2FoldChange), !is.na(padj)) %>% 
  rownames_to_column(var = "ENSEMBL")

# Add column 'regulation'
Untreated_vs_IsoCTR_DEGs$regulation <-
ifelse( Untreated_vs_IsoCTR_DEGs$padj < 0.05 &
          Untreated_vs_IsoCTR_DEGs$log2FoldChange > 1, "Up",
  ifelse( Untreated_vs_IsoCTR_DEGs$padj < 0.05 &
            Untreated_vs_IsoCTR_DEGs$log2FoldChange < -1, "Down",
          "NotSig")
)

# Write the file
write.table(Untreated_vs_IsoCTR_DEGs,
            file = "./projects/project_1/DEG_List_backup/Untreated_vs_IsoCTR_DEGs.txt",
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)
```

**How many genes were (significantly) up/down-regulated in 'Treated' compared to 'Untreated'?**

```{r echo=FALSE, message=FALSE, warning=FALSE}
Treated_vs_Untreated <- 
  results(dds_filt, 
          contrast=c("sample_type","Treated","Untreated"),
          alpha = 0.05)
summary(Treated_vs_Untreated)
```

# Crossing DEG sets: Venn diagram

We want to intersect 3 sets of DEGs:

-   **Untreated vs IsoCTR:** genes that are differentially expressed in DMD organoid compared to isogenic control

-   **Treated vs Untreated:** genes that are differentially expressed in AAV-treated DMD organoid compared to non-treated (presumably 'corrected' by gene therapy)

-   **Treated vs IsoCTR:** presumably 'not corrected' by gene therapy

| Note: We can use `abs(log2FoldChange) > 1` but it seems too strict.

```{r venn_diagram, echo=TRUE, fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
x <- list(
  rownames(Untreated_vs_IsoCTR %>% 
             as.data.frame() %>% 
             filter(padj < 0.05 & abs(log2FoldChange) > 0)),
  rownames(Treated_vs_Untreated %>% 
             as.data.frame() %>% 
             filter(padj < 0.05 & abs(log2FoldChange) > 0)),
  rownames(Treated_vs_IsoCTR %>% 
             as.data.frame() %>% 
             filter(padj < 0.05 & abs(log2FoldChange) > 0))
)

library("ggVennDiagram")
ggVennDiagram(x, label_alpha = 0,
              label = "count",
              category.names = c("Untreated\nvs IsoCTR",
                                 "Treated\nvs Untreated",
                                 "Treated\nvs IsoCTR")) +
  
  theme(plot.margin = margin(1, 4, 1, 4, "cm"),
        text = element_text(size = 14)) +
  
  scale_fill_gradient(low = "white", high = "orange")
```

# Gene Set Enrichment Analysis (GSEA)

Prior to GSEA, we need to create a **ranked list** of the genes.

<details>

<summary>Tidy and save Treated_vs_Untreated DEG table</summary>

```{r echo=TRUE}

# Remove genes where log2FC or padj is missing
# Make row names a column EMSEMBL

Treated_vs_Untreated_DEGs <- 
  Treated_vs_Untreated %>%
  as.data.frame() %>% 
  filter(!is.na(log2FoldChange), !is.na(padj)) %>% 
  rownames_to_column(var = "ENSEMBL")

# Add column 'regulation'
Treated_vs_Untreated_DEGs$regulation <-
ifelse( Treated_vs_Untreated_DEGs$padj < 0.05 &
          Treated_vs_Untreated_DEGs$log2FoldChange > 1, "Up",
  ifelse( Treated_vs_Untreated_DEGs$padj < 0.05 &
            Treated_vs_Untreated_DEGs$log2FoldChange < -1, "Down",
          "NotSig")
)

# Write the file
write.table(Treated_vs_Untreated_DEGs,
            file = "./projects/project_1/DEG_List_backup/Treated_vs_Untreated_DEGs.txt",
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)
```

</details>

A bunch of ranking metrics can be used to rank the list of input genes. In the end, they should all correlate to the two phenotypes signatured by the gene set that we're interested in.

Here we'll try using the `log2FoldChange` in DESeq2.

| Using shrinkage of LFC estimates (toward 0) is to avoid inflation by genes with small counts/large standard errors.

```{r rank_list, echo=TRUE, message=FALSE, warning=FALSE}
Treat_vs_Untreated_DEGs <-
read.table(file = "./projects/project_1/DEG_List/Treated_vs_Untreated_DEGs.txt",
           header = TRUE, sep = "\t")

Treat_vs_Untreated_DEGs_ranked <- Treat_vs_Untreated_DEGs %>% 
  arrange(desc(log2FoldChange)) %>%
  pull(log2FoldChange, name=ENSEMBL) # create named vector

head(Treat_vs_Untreated_DEGs_ranked)
```

The **gene set** of interest can be obtained via the [Molecular Signatures Database (MSigDB)](https://www.gsea-msigdb.org/gsea/msigdb).

```{r get_gene_set, echo=TRUE, message=FALSE, warning=FALSE}
library(msigdbr)
GOBP_all <- msigdbr(
  species = "Homo sapiens",
  collection = "C5",
  subcollection = "GO:BP"
) %>% 
  dplyr::select(gs_name, ensembl_gene)
head(GOBP_all)
```

Running GSEA with all GOBP gene sets:

```{r run_gsea, echo=TRUE, message=FALSE, warning=FALSE}
library(clusterProfiler)
gsea_GOBP_all <- GSEA(
  Treat_vs_Untreated_DEGs_ranked,
  TERM2GENE = GOBP_all,
  pvalueCutoff = 0.05)

# List of gene sets with significant enrichment
head(gsea_GOBP_all@result, 10)$ID
```

```{r dotplot, echo=TRUE, fig.height=10, fig.width=8, message=FALSE, warning=FALSE}
library(enrichplot)
dotplot(gsea_GOBP_all, showCategory = 20,
        color = "NES") +
  ggtitle("Top 20 enriched pathways")
```

<details>

<summary>**How to see only terms that relate to "inflammation"?**</summary>

```{r inflamm_GOBP, echo=TRUE, fig.height=5, fig.width=8, message=FALSE, warning=FALSE}
# Subset result table for terms containing "*inflam*"
inflam_terms <- gsea_GOBP_all@result %>%
  filter(grepl("inflam", Description, ignore.case = TRUE))

# Create a new gseaResult object
gsea_GOBP_inflam <- gsea_GOBP_all
# The result table in this new object is the subset one
gsea_GOBP_inflam@result <- inflam_terms

# Plot only '*inflam*'-related sets
dotplot(gsea_GOBP_inflam, color = "NES",
        showCategory = nrow(inflam_terms)) +
  ggtitle("Inflammation-related GO terms")
```

</details>

Plot enrichment score for terms/signatures that we're particularly interested in:

```{r gsea_plot, echo=TRUE, fig.height=8, fig.width=6, message=FALSE, warning=FALSE}
# Visualization
library(patchwork)
p1 <- gseaplot2(gsea_GOBP_all, geneSetID = "GOBP_POSITIVE_REGULATION_OF_INFLAMMATORY_RESPONSE",
                            pvalue_table = TRUE)
p1_list <- wrap_plots(p1, ncol = 1,
                      heights = c(1,0.2,0.5),
                      align = "v")

p2 <- gseaplot2(gsea_GOBP_all, geneSetID = "GOBP_MICROTUBULE_ORGANIZING_CENTER_ORGANIZATION",
                            pvalue_table = TRUE)
p2_list <- wrap_plots(p2, ncol = 1,
                      heights = c(1,0.2,0.5),
                      align = "v")

# Combine 2 plots
library(cowplot)
plot_grid(p1_list, p2_list, ncol = 1, align = "v")
```

# Convert ENSEMBL ID to gene symbol (optional)

There are many ways to get gene names based on Ensembl IDs. However, some IDs, usually corresponding to non-coding RNAs, do not match a gene symbol (from [HGNC](https://www.genenames.org/) database). If we care only about protein-coding genes, maybe go back to `featureCounts` and consider reads that map CDS.

-   Here we'll try keeping the IDs as they are for those that don't have a gene symbol.

-   Same for IDs that match multiple gene symbols, we keep the Ensembl IDs to make it less ambiguous.

<details>

<summary>**Convert EMSEMBL ID to gene symbol**</summary>

```{r convert_to_gene_symbol, echo=TRUE, message=FALSE, warning=FALSE}

library(org.Hs.eg.db)
annots <- AnnotationDbi::select(org.Hs.eg.db,
                 
                 # ensembl ids in count table
                 keys=rownames(count_tab),
                 
                 columns=c("SYMBOL", "GENENAME"), 
                 keytype="ENSEMBL")

# Replace missing SYMBOL with the ENSEMBL ID
annots$SYMBOL[is.na(annots$SYMBOL)] <- annots$ENSEMBL

# Collapse SYMBOL, but if >1 SYMBOL, use the ENSEMBL ID instead
annots <- annots %>%
  group_by(ENSEMBL) %>%
  summarise(
    SYMBOL = ifelse(n_distinct(SYMBOL) == 1,
                    unique(SYMBOL),
                    ENSEMBL)
  )

# Add 'SYMBOL' column to DEG table
Treated_vs_Untreated_DEGs_annot <- Treated_vs_Untreated_DEGs %>% left_join(annots, by = "ENSEMBL")
```

</details>

```{r echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(head(Treated_vs_Untreated_DEGs_annot), caption = "Treated_vs_Untreated DEG table with gene symbols") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

**Tutorials**

[Analyzing RNA-seq data with DESeq2](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)

**References**
